# Likelihoodism {#likelihoodism}

* The likelihood function is all you need and is interpreted as evidence for a statistical hypothesis (law of likelihood)
* Statistical hypotheses are compared by the ratio of their likelihoods.
* A [statistical modeling mindset](#statistical-modeling) with [frequentism](#frequent-inference) and [Bayesianism](#bayesian-inference) as alternatives.

<!-- metaphor -->
A frequentist, a Bayesian, and a likelihoodist walk into a bar, a wine bar.
The sommelier quickly joins the three. 
The Bayesian wants to hear the sommelier's opinion first before trying the wines.. 
The frequentist asks the sommelier about the tasting process: is the number of wines fixed in advance? Is the tasting over when the customer has found a suitable wine? How are subsequent wines selected?
The likelihoodist politely tells the sommelier to fuck off.

<!-- problem with frequentism -->
[Frequentist inference](#frequentist-inference) has a long list of limitations.
But it's still the dominant statistical mindset in science and elsewhere.
Bayesian analysis has seen a resurgence thanks to increased computational power for sampling from posteriors with MCMC.
But using subjective prior probabilities doesn't sit well with many statisticians.
Could there be another way to "reform" the frequentist mindset?
A mindset without the flawed hypothesis testing and without priors?

Welcome to the **likelihoodist mindset.**

<!-- underdog -->
I studied statistics for 5 years, worked as a statistician and data scientist for 3 years, and then did PhD studies in machine learning for 4.5 years.
In those 12 years of statistics, I never learned anything about likelihoodism.
It's fair to say that likelihoodism is the underdog.
Likelihoodism leads a shadowy existence while Bayesianism and frequentism are engaged in an epic battle.

<!-- why learn about it -->
Likelihoodism is the purist among the statistical modeling mindsets.
A mindset that focuses entirely on the likelihood function.
Likelihoodism is an attempt to make statistics as objective as possible.

<!-- short reminder: likelihood -->
**All three mindsets use likelihood functions in different ways.**
A quick recap: The likelihood function is the same as the data density function, but the roles of data and parameters are reversed.
Data $X$ are "fixed" and the likelihood is a function of the parameters $\theta$ $P(\theta; X) = P(X = x | \theta)$.
The likelihood links observed data to theoretic distributions.
<!-- use of the likelihood function -->
Bayesians multiply prior distributions with the likelihood to get the posterior distributions of the parameters.
Frequentists use the likelihood to estimate parameters and construct "imagined" experiments that teach us about long-run frequencies (hypothesis tests and confidence intervals).
Likelihoodists view the likelihood as evidence derived from data for a statistical hypothesis. 
Likelihoodists emphasize the likelihood and reject the non-likelihood elements from frequentism and Bayesianism:
Likelihoodists reject priors because they are subjective;
Likelihoodists reject the frequentists' reliance on "imagined" experiments because these never-observed experiments violate the likelihood principle.

```{r three-stat-mindsets, fig.cap = "How Bayesianism, frequentism, and likelihoodism overlap and differ. Figure inspired by [Greg Gandenberger](https://gandenberger.org/2014/07/28/intro-to-statistical-methods-2/)."}
library(ggplot2)
circleFun <- function(center = c(0,0),diameter = 1, npoints = 100){
  r = diameter / 2
    tt <- seq(0,2*pi,length.out = npoints)
    xx <- center[1] + r * cos(tt)
    yy <- center[2] + r * sin(tt)
    return(data.frame(x = xx, y = yy))
}

circle1 = circleFun(c(0.7, -1), 2.3)
circle2 = circleFun(c(2.3, -1), 2.3)
circle3 = circleFun(c(1.5, 0.3), 2.3)
csize = 1.5
#geom <- path will do open circles, geom <- polygon will do filled circles
ggplot(mapping = aes(x,y)) +
  geom_path(data = circle1, size = csize) +
  geom_path(data = circle2, size = csize) + 
  geom_path(data = circle3, size = csize) + 
  annotate("text", label = "No priors", x = 0.5, y = -1, size = 10) +
  annotate("text", label = "Provides \n belief/action", x = 2.6, y = -1, size = 10) +
  annotate("text", label = "Likelihood \n Principle", x = 1.5, y = 0.5, size = 10) +
  annotate("text", label = "Bayesianism", x = 2.1, y = -0.2, size = 6, angle = 40) +
  annotate("text", label = "Likelihoodism", x = 0.85, y = -0.25, size = 6, angle = -40) +
  annotate("text", label = "Frequentism", x = 1.5, y = -1.3, size = 6, angle = 90) +
  coord_fixed() +
  theme_void()

```

<!-- likelihood principle as foundation -->
But what is the likelihood principle that is so central to likelihoodism?

## Likelihood Principle

"The likelihood principle asserts that two observations that generate identical likelihood functions are equivalent as evidence." [@richard2017statistical]
As a consequence, all evidence that comes from the data about a quantity of interest $\theta$ has to be part of the likelihood function $P(\theta;X)$.
If we reverse the statement: If information from the data influences the analysis but is not part of the likelihood, then the likelihood principle is violated.

Let's say we want to estimate the average waiting time for a public bus.
To model waiting times, the exponential distribution is a good choice.
So we could assume that $X \sim Exp(\lambda)$, where $\lambda$ can be interpreted as the inverse waiting time.
The expected waiting time is $\frac{1}{\lambda}$.
We have collected $n$ waiting times $x_1, \ldots, x_n$.
The likelihood function is: 

$$P(\lambda; x_1, \ldots, x_n) = \lambda^n exp\left(-\lambda \sum_{i=n}^n x^{(i)}\right).$$

In all three mindsets, we could work with this likelihood.
Bayesians would, in addition, assume a prior distribution for $\lambda$.
Whether the likelihood principle is violated depends on what we do after calculating the likelihood.
Bayesians obtain a posterior distribution for $\lambda$ that is interpreted as belief about the parameter. 
The likelihoodist might report the likelihood region for $\lambda$, which can be interpreted as relative evidence for a range of parameter values compared to the maximum likelihood estimate for $\lambda$.
Both the Bayesian and the likelihoodist approaches adhere to the likelihood principle:
All evidence from the data about $\lambda$ is included in the likelihood.
Bayesians use priors, but as long as they don't include any information from the data, it's fine. [^empirical-priors]

The frequentist might test whether $\lambda$ is significantly smaller than a certain value. 
When performing such a test, the frequentist has to "imagine" experiments under the null hypothesis distribution.
But the null hypothesis is not part of the likelihood.
Frequentists choose the distribution under the null hypothesis based on how they "imagine" repetitions of the sample or experiment.
This, in turn, depends on how the experiment was conducted or how the data were collected in the first place.
You will see later an example of a coin toss where the same data from different experiments lead to  different conclusions in the frequentist mindset.

<!--
The likelihood principle is a consequence of 1) the sufficiency principle and 2) the conditionality principle.[@birnbaum1962foundations]
The sufficiency principle says that a sufficient statistic $S(X)$ of a quantity of interest summarizes all relevant evidence from the data.
The conditionality principles says that experiments or samples that were not performed should be ignored.
Many frequentist approaches directly violate the conditionality principle.
Statistical hypothesis tests are designed around experiments that never happen.
-->

So the big difference between frequentism and likelihoodism is the likelihood principle.
The likelihood principle gives the likelihood function the monopoly over data evidence.
But the likelihood principle alone is not sufficient to create a coherent modeling mindset.
We need the law of likelihood.

## Law of Likelihood

The law of likelihood is the foundation for using the likelihood function alone formaking modeling decisions.
The law of likelihood says [@hacking1965logic]:

* Given:
  * Hypotheses $H_1$ and $H_2$; data $\mathbf{x} = \{x^{(1)}, \dots, x^{(n)}\}$.
  * Likelihood for $H_1$ is larger than for $H_2$: $P_{H_1}(X = \mathbf{x}) > P_{H_2}(X = \mathbf{x})$.
* Then:
  * Observation $X=\mathbf{x}$ is evidence supporting $H_1$ over $H_2$.
  * The likelihood ratio $P_{H_1}(x)/P_{H_2}(x)$ measures the strength of this evidence.

The hypotheses can be the same statistical model, but with different parameter values $\theta$.
Returning to the bus waiting time example, $H_1$ could be that $\lambda = 1$, and $H_2$ could be that $\lambda = 0.5$.
The resulting likelihood ratio might be:

$$P(\lambda = 1;x_1, \ldots, x_n)/P(\lambda = 0.5;x_1, \ldots, x_n) = 4$$.

The likelihood ratio is the likelihood of one statistical hypothesis divided by the likelihood of another.
As a reminder, statistical hypotheses are statistical models where, optionally, some or all of the parameters are assigned by hand rather than learned from the data.
The law of likelihood tells us, that to compare hypotheses $H_1$ and $H_2$ with their likelihood ratio:

$$\frac{P(H_1;X = \mathbf{x})}{P(H_2;X = \mathbf{x})}$$

In frequentism, likelihood ratios are often used as test statistics for hypothesis tests.
In likelihoodism, the likelihood ratio is interpreted as evidence.

Likelihoodists may use a rule of thumb for judging the strength of evidence.
For example, a likelihood ratio of 8 is considered fairly strong and 32 or more is considered "strong favoring".[@richard2017statistical]
In our example, a likelihood ratio of 4 in favor of $H_1: \lambda = 1$ over $H_2: \lambda = 0.5$ is not enough to be "fairly strong".
$H_1$ and $H_2$ can also be more complex hypotheses, such as regression models with different covariates or assumptions.

<!-- stronger law -->
**The law of likelihood is stronger than the likelihood principle:**
The likelihood principle states that the all evidence from the data must be in the likelihood;
**The law of likelihood describes how evidence can be quantified and compared.**
And this is where Bayesian inference and likelihoodism differ.
Bayesians are not guided by the law of likelihood, but by Bayesian updating and a subjective interpretation of interpretability.

<!-- clear vision -->
The law of likelihood makes it clear how we can compare statistical hypotheses:
Not by hypothesis testing, but by their likelihood ratios.
The larger the ratio, the stronger the evidence for one hypothesis over another.

<!-- shortcomings -->
This is also where likelihoodism reaches a dead end.
The ratio may only be interpreted as evidential favoring.
The likelihoodist mindset doesn't come with guidance on what we should believe about the parameters or what decision/action to take based on the results.
The likelihood ratio only tell us which hypothesis is favored.


## Likelihood Intervals

<!-- using likelihood intervals -->
Likelihood intervals are the likelihoodist analogue to frequentist confidence interval and Bayesian credible intervals.
Likelihood intervals are interpreted in terms of, you guessed it, relative likelihood.
The likelihood interval of a model parameter $\theta$ is the set of all $\theta$ values that yield a relative likelihood greater than a certain threshold:

$$\left\{\theta: \frac{L(\theta| X)}{L(\hat{\theta}| X)} \geq \frac{p}{100}\right\}$$

The $\hat{\theta}$ is the optimal $\theta$ after fitting the model using maximum likelihood estimation or another optimization method.
Let's say for a logistic regression model coefficient $\beta_j$: $\hat{\beta}_j = 1.1$. <!-- fix vim_ -->
Then an interval could be $[0.9; 1.3]$.
The role of the constant $p$ is similar to the one of the $\alpha$-level for confidence and credible intervals:
It specifies the size of the interval.
See figure \@ref(fig:likelihood-interval).
Each $\theta$-value within that interval can be seen as constituting a different hypothesis.
And these hypotheses are compared with the optimal model $\theta = \hat{\theta}$.  

```{r likelihood-interval, fig.cap = "1/4 and 1/16 likelihood ratio intervals.", fig.height = 3, fig.width = 10}
n = 1000
library(ggplot2)

x = seq(from = -3, to = 3, length.out = n)
dat = data.frame(x = x, dens = dnorm(x, sd = 0.8))
dat$x = dat$x
dat$lr = dnorm(0) / dat$dens
lr4l = min(dat$x[dat$lr <= 4])
lr4u = max(dat$x[dat$lr <= 4])

# y coordinates for these segments
lr4ly = dat$dens[dat$x == lr4l]
lr4uy = dat$dens[dat$x == lr4u]

lr16l = min(dat$x[dat$lr <= 16])
lr16u = max(dat$x[dat$lr <= 16])

# y coordinates for these segments
lr16ly = dat$dens[dat$x == lr16l]
lr16uy = dat$dens[dat$x == lr16u]

ggplot(dat) +
  geom_line(aes(x = x, y = dens)) +
  annotate("segment", x = lr4l, xend = lr4u, y = lr4ly, yend = lr4uy, arrow = arrow(ends = "both", angle = 90)) +
  annotate("segment", x = lr16l, xend = lr16u, y = lr16ly, yend = lr16uy, arrow = arrow(ends = "both", angle = 90)) +
  annotate("point", x = 0, y = max(dat$dens), size = 3) +
  annotate("label", label = "LR 4", x = lr4l -0.2, y = lr4ly) +
  annotate("label", label = "LR 16", x = lr16l -0.2, y = lr16ly) +
  theme_void()

```



## Why Frequentism Violates the Likelihood Principle

Many frequentist methods violate the likelihood principle because they require "imagined" experiments.
But frequentists need these theoretic distribution to compute p-values, for example.
Statistical null hypothesis tests also depend on stopping criteria for data collection, which is in conflict with the likelihood principle.

<!-- motivation for example -->
The following example shows how frequentism violates the likelihood principle because of different stopping criteria.
<!-- starting coin example -->
Suppose we have a coin.
We want to find out whether it's fair, or whether maybe head turns up too often.
I still don't know why statistician's are so upset about unfair coins, but that's the way it is.
We call $\theta$ the probability of head.
We have two hypotheses:

$$H_0: \theta = 0.5 \text{ and } H_1: \theta > 0.5$$

$H_0$ means that the coin is fair.
$H_1$ claims that head comes up more often than tail.
We define two random variables: the number of heads $X$, and the number of coin tosses $Y$.

<!-- two types of experiments -->
We perform two experiments with a different setup but with the same results:

1. Toss the coin 12 times. We observe head 9 out of 12 times.
1. Toss the coin until tail is observed 3 times. The third tail appears on the 12th toss.

Both experiments have the exact same outcome, but we defined the stopping criteria for the experiments differently:
In experiment 1), we fixed the number of tosses $Y$.
In experiment 2), we fixed the number of heads $X$.
Should we reach different conclusions about the fairness of the coin?
What do you think?

<!-- likelihoodist view of the experiment -->
Both experiments give the same likelihood, up to a constant factor. Experiment 1):

$$L(\theta | X = 3) = \binom{12}{3} \theta^3 (1 - \theta)^9  = 220 \theta^3 (1 - \theta)^9 $$

And experiment 2):

<!-- the 2 out of 11 results from the fact that the last toss must be a head -->
$$L(\theta | Y = 12) = \binom{11}{2} \theta^3 (1 - \theta)^9  = 55 \theta^3 (1 - \theta)^9$$

So the likelihoodists say that both experiments carry the same evidence.
The likelihood intervals would be the same for both experiments.

<!-- Frequentist view on the experiments -->
Frequentists would come to different conclusions depending on the experiment.
Frequentists include results that have not occurred but depend on how the experiments are conducted.
They assume that $H_0$ is true and infer how the test statistic is distributed in future experiments under $H_0$.
Then frequentists place the estimated value of $\hat{\theta}$ within this distribution of imagined experiments and see how extreme the result is.
In experiment 1), where the number of tosses is fixed, experiment outcomes of 9, 10, 11, or 12 heads are more extreme than the actual experiment outcome.
In the other experiment, the number of tails is fixed.
More extreme outcomes in experiment 2) are all possible experiments where we observe more than 12 tosses.
This includes, for example, the  experiment where the third tail only comes up after 1108 tosses.

When we test $H_0$ vs. $H_1$ in experiment 1), we get:

$$P_{H_0}(X \geq 9) = \sum_{x = 9}^{12} \binom{x}{12} 0.5^x (1 - 0.5)^{12 - x} = 0.073$$

At a significance level of $\alpha = 0.05$, we would not reject the fair coin hypothesis.

For experiment 2), we assume a negative binomial distribution:

$$P_{H_0}(Y \geq 12) = \sum_{y=12}^{\infty}\binom{3 + y - 1}{2} 0.5^y 0.5^3 = 0.0327$$

The p-value is now smaller than 0.05, and with that the coin is significantly unfair.

In frequentist inference, the way data are collected and the way experiments are designed affect the results.

This has much more subtle consequences than I've illustrated so far.
Imagine a domain expert asks you to perform an analysis with 1000 data points.
As a frequentist, you need to know how those 1000 data points were collected.
What was the stopping criterion for data collection?
If the domain expert only planned to collect 1000 data points, that's fine.
But if the expert says she would collect more data  depending on the outcome of the analysis, then that changes the analysis, which is a violation of the likelihood principle.

## Strengths

* Likelihoodism inherits all the strengths of statistical models.
* It's a coherent modeling approach: all information is contained in the likelihood. Frequentism, in contrast, is more fragmented with long lists of differently motivated statistical tests and confidence intervals.
* Like Bayesian inference, likelihoodist inference is also consistent with the likelihood principle. Therefore neither is affected by experimental design, as is the case with frequentism. 
* Likelihoodism is arguably the most objective of the statistical modeling mindsets. No priors, no imagined experiments.
* Likelihoodist ideas can improve the reporting of Bayesian results. For example, Bayesians can additionally report likelihood ratios as evidence.
* A significance test might reject $H_0$, even if the evidence for $H_0$ is greater than for $H_1$. Likelihoodism doesn't have this problem.

## Limitations 

* Likelihoodism doesn't provide guidance in the form of belief or decision. Evidence is less practical, and the statistician has no certainty about which the final model is and how to work with it. This is the strongest argument against likelihoodism, and maybe the reason why we don't see it in practice.
* To be more specific: There is no theoretical underpinning for saying when there is enough evidence to choose one hypothesis over another.
* With a likelihoodist mindset, we can can only compare simple hypotheses where all parameters are specified. Composite hypotheses for ranges of parameters are impossible. Likelihoodism can't compare  $\theta > 0$ versus $\theta \leq 0$. Only, for example $\theta = 1$ against $\theta = 0$.
* Likelihoodism allows only relative statements. It can't state the probability that a statistical hypothesis is true -- only how its evidence compares to another hypothesis.

## Resources

* The book "Statistical Evidence: A Likelihood Paradigm" is good introduction to likelihoodism, if you have some background as a statistician.
* I've found the [Greg Gandenberger's blog](http://gandenberger.org/) super helpful in learning about likelihoodism. He takes a more philosophical viewpoint and argues against likelihoodism and for Bayesianism. This critique is most detailed in his essay "Why I am not a likelihoodist".[@gandenberger2016not] 


[^empirical-priors]: The likelihood principle is violated when data is used to inform the prior. For example, empirical priors which make use of the data violate the likelihood principle.
